# 6. Analysis of longitudinal data

## 1. RATS data

### a) Read in data:
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
RATS <- read.csv("data/rats.csv", row.names = 1, header= TRUE)
RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
summary(RATS)
str(RATS)
```
This is data from a nutrition study conducted on three groups of rats. The groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately) weekly, except in week seven when two recordings were taken) over a 9-week period
. Let's create a graphical representation to understand the data better:

```{r}
p1 <- ggplot(RATS, aes(x = Time, y = Weight, linetype = ID))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + facet_grid(. ~ Group, labeller = label_both)
p4 <- p3 + theme_bw() + theme(legend.position = "none")
p5 <- p4 + theme(panel.grid.minor.y = element_blank())
p6 <- p5 + scale_y_continuous(limits = c(min(RATS$Weight),max(RATS$Weight)))
p6
```
A visual assessment of the graphs shows, that rats in group 1 with weight < 300 did not change their weight dramatically, there is possibly a slight decrease. However, individuals with higher weight in the beginning of the study, especially in group 2 seemed to have increased their weight over time.
In order to make a general assessment of weight changes we will summarize the rats according to their group.

## Summarize Groups
```{r}
n <- RATS$Weight %>% unique() %>% length()

RATSS <- RATS %>% group_by(Group, Time) %>% summarise( mean=mean(Weight), se=sd(Weight)/sqrt(n)) %>% ungroup()
glimpse(RATSS)

p1 <- ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group))
p2 <- p1 + geom_line() + scale_linetype_manual(values = rep(1:10, times=4))
p3 <- p2 + geom_point(size=1) + scale_shape_manual(values = rep(1:10, times=4))
p4 <- p3 + geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3)
p5 <- p4 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6 <- p5 + scale_y_continuous(name = "mean(Weight) +/- se(Weight)")
p6
```
The graphical summary of the data shows that within the groups the weight varies only by little. Overtime, group 1 has almost no increase, while group 2 and 3 increase more significantly. Group 2 varies the most, in fact it is the only data for which the errorbars are easy to spot by eye.
Step by step we can condense the data further into more concise graphs:

```{r}
p1 <- ggplot(RATS, aes(x = factor(Time), y = Weight, fill = Group))
p2 <- p1 + geom_boxplot(position = position_dodge(width = 0.9))
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + scale_x_discrete(name = "Time (days)")
p4
```
The rows of boxplots show that all groups contain outliers, or in this case one rat is an outlier for each groups. Nevertheless, it once more confirms earlier suspicions and so does the following even simpler graph:

```{r}
# Make a summary data of the post treatment weeks (1-8)
RATS8S <- RATS %>% filter(Time > 0) %>% group_by(Group, ID) %>% summarise( mean=mean(Weight) ) %>% ungroup()
glimpse(RATS8S)
p1 <- ggplot(RATS8S, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(bprs), weeks 1-8")
p5
```
So lets see what happens if we remove these outliers:

```{r}
RATS8S1 <- RATS8S %>% filter( (mean > 240 & Group == 1) | (mean < 480 & Group == 2) | (mean > 500 & Group == 3) )
glimpse(RATS8S1)
p1 <- ggplot(RATS8S1, aes(x = Group, y = mean))
p2 <- p1 + geom_boxplot()
p3 <- p2 + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4 <- p3 + stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white")
p5 <- p4 + scale_y_continuous(name = "mean(Weight)")
p5
```
After removal of the outliers it becomes obvious that all the variance in Group 2 was basically caused by this one outlier. With a precise summary of our data at hand we can go ahead and see if we can fit a model onto it:


```{r}
# For establishing a baseline raw data is needed
RATSRAW <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", stringsAsFactors = F, sep =  "\t")
# Once again remove outliers
RATSRAW <- RATSRAW  %>% filter( (! ID == 2) & (! ID == 12) & (! ID == 13) )
baseline <- RATSRAW$WD1
RATS8S2 <- RATS8S1 %>%
  mutate(baseline)
# Fit the ANCOVA model and see the results:
fit <- lm(mean ~ baseline + Group, data = RATS8S2)
summary(fit)
anova(fit)
```
E voilá, **finally** we have a model with very good significance!